Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-11-19 manuscript/manuscript.tex
sorted and reformatted bib

The impact of NYC 311 data extends beyond operational efficiency; it
has become instrumental in shaping City governance and community
engagement. Open data not only ensures governmental transparency
but also empowers researchers, civic developers, and the general
public. The data has been pivotal in providing advice on shelters
during emergencies, handling inquiries during the COVID\mbox{-}19 pandemic,
enforcing standards between landlords and tenants, reallocating taxi
routes based on analyses by the Taxi and Limousine Commission (TLC),
and improving responsiveness across City Agencies.
\jy{need to expand the importance of 311 SR data to the research
  community: such as
  resource allocation/response \citep{zha2014profiling, raj2021swift},
  social equity \citep{kontokosta2021bias},
  street flood \citep{agonafir2022understanding}, etc.}

Our investigation utilizes a 311 SR dataset covering  2\mbox{-}years (2022-2023).
This sample was collected on 15 September 2024, and is used 
to conduct detailed analysis in areas such as data 
consistency, data validity, redundant fields, etc.. Instructions 
on downloading this file from the NYC Open Data Portal 
are available in the supplemental material. Here are some 
characteristics of the 311 SR dataset used in this analysis:

\begin{itemize}[left=1.5em]
	\item The dataset is approximately 3.7 GB in size and contains 
	over 6.4 million rows. It was exported as a CSV file.

@@

	\item In addition to  \texttt{incident\_address}, there are five additional location fields:

	\texttt{latitude \& longitude}, \texttt{street\_name}, \texttt{landmark}, USGS State Plane 
	Coordinate System, and \texttt{block} \textbf{\#}.

@@

During the 10\mbox{-}year period from 2014 to 2023,  NYC's 311 system
experienced a 50\% increase in the number of SRs, 
reflecting a growing public reliance on and usage of 
the 311 system. (During this 10\mbox{-}year time-frame, the 
population of NYC grew just 0.6\%.). The most significant increase
in activity occurred in 2020, when the volume of requests 
surged to 3.23 million during the COVID\mbox{-}19 pandemic, 
illustrating the essential role of the 311 system in supporting 
NYC residents during times of crisis. This steady 
rise in SRs can be at least partially attributed to the increased 
accessibility of the 311 system via online and mobile 
platforms, as well as heightened public awareness of the service. 
While the spike in 2020 was exceptional, the broader trend 
indicates sustained growth in SR volume in future years,  
highlighting the system's expanding role in managing 
both routine City operations and extraordinary events.

Figure~\ref{fig:SRcountbyAgency} provides a breakdown of SRs by
the responsible Agency, showing the cumulative 
percentage of SRs handled by each
agency over the 2\mbox{-}year time-frame. Note that the
distribution of SRs is heavily concentrated among a few key 
agencies. The six largest agencies are:

\begin{enumerate}[left=1.5em]
    \item New York Police Department (NYPD): 43\% of all SRs
    \item Housing Preservation and Development (HPD): 21\%
    \item New York City Department of Sanitation (DSNY): 10\%
    \item Department of Transportation (DOT): 7\%
    \item Department of Environmental Protection (DEP): 5\%
    \item Department of Parks and Recreation (DPR): 4\%
\end{enumerate}

@@

The \textit{Mostly Empty} category includes fields such as
\texttt{taxi\_company\_borough}, \texttt{due\_date},
\texttt{road\_ramp}, and \texttt{bridge\_highway\_name}.
The \textit{Partially Empty} category includes fields such as
\texttt{location\_type}, \texttt{landmark}, 
and \texttt{cross\_street\_1 \& \_2}. And the \textit{Few/None Empty} 
category includes \texttt{created\_date}, \texttt{complaint\_type},
\texttt{agency}, and \texttt{status}. For some analysis efforts, it may be
prudent to inquire further as to why some fields are 
mostly or almost always blank. Figure~\ref{fig:blank_fields} is 
a graphic depiction of total empty (blank \& N/As) for each 
field illustrating the grouping into 
the \textbf{Most}, \textbf{Partial}, and \textbf{Few/None} categories.

@@

\begin{itemize}[left=1.5em]
    \item SRs with a \texttt{closed\_date} that occurs before the 
    \texttt{created\_date} resulting a nonsensical ``negative duration''. (12,450)
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} in 
    the far distant past, e.g. ``1900-01-01''. (8)
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} that 
    match \textbf{to the second} creating an impossible ``zero duration''. (163,720)
    \item An unusually large number of SRs closed and/or created 
    exactly at midnight, to the second. (247,000)
\end{itemize}

@@

\begin{table}[tbp]
  \centering
  \caption{Sample  Negative Durations [excluding extreme values that are <$-$730 days]}
   \label{tab:largest-errors}
 	\begin{tabular}{l l l r l}
    \toprule
    {created\_date} & {closed\_date} & {duration (days)}  & {agency} \\
    \midrule
    2023-01-27 14:40:00 & 2022-01-14 14:40:00 & $-$378 & DOT \\
    2023-01-18 10:06:00 & 2022-01-12 10:06:00 & $-$371 & DOT \\
    2023-01-27 14:36:00 & 2022-01-22 14:35:00 & $-$370 & DOT \\
    2023-01-11 11:10:00 & 2022-01-09 11:10:00 & $-$367 & DOT \\
    2023-12-18 03:13:00 & 2023-01-16 13:10:00 & $-$335 & DOT \\
    \bottomrule
    \end{tabular}
 \end{table}

\paragraph{Identical Created \& Closed Dates (Zero Durations):}
A more prevalent issue occurs when the \texttt{closed\_date} and 
\texttt{created\_date} have exactly the same date\mbox{-}time stamp, 
to the second. This creates a \textbf{zero duration} for the 
SR, which is nonsensical. There are 163,720 such zero duration 
SRs, representing 2.6\% of all non\mbox{-}blank data. 
Additionally, 99\% percent of these zero duration SRs 
occur within five agencies: Department of Health \& Mental 
Hygiene (DHMH), DOT, Department of Buildings (DOB), DSNY, and 
DEP. This distribution \textbf{does not} 
mirror the overall SR distribution by agency, suggesting 
an Agency\mbox{-}specific issue.

>>>>

The impact of NYC 311 data extends beyond operational efficiency; it
has become instrumental in shaping City governance and community
engagement. Open data not only ensures governmental transparency
but also empowers researchers, civic developers, and the general
\jy{back up with some references:
  \citep{minkoff2016nyc, o2017uncharted, kontokosta2021bias}}
public. The data has been pivotal in providing advice on shelters
during emergencies, handling inquiries during the COVID\mbox{-}19 pandemic,
enforcing standards between landlords and tenants, reallocating taxi
routes based on analyses by the Taxi and Limousine Commission (TLC),
and improving responsiveness across City Agencies.
\jy{need to expand the importance of 311 SR data to the research
  community: such as
  resource allocation/response \citep{zha2014profiling, raj2021swift},
  social equity \citep{white2018promises, kontokosta2021bias},
  noise \citep{dove2022sounds},
  street flood \citep{agonafir2022understanding}, etc.}

Our investigation utilizes a 311 SR dataset covering  2\mbox{-}years (2022-2023).
This sample was collected on 15 September 2024, and is used 
to conduct detailed analysis in areas such as data 
consistency, data validity, redundant fields, etc.. Instructions 
on downloading this file from the NYC Open Data Portal 
are available in the supplemental material. Here are some 
characteristics of the 311 SR dataset used in this analysis:
\jy{need to clarify that we've made the column names to lower cases
  and replaced spaces with underscores.}
\begin{itemize}[left=1.5em]
\item The dataset is approximately 3.7 GB in size and contains 
  over 6.4 million rows. It was exported as a CSV file.

@@

\item In addition to  \texttt{incident\_address}, there are five additional location fields: 
  \texttt{latitude \& longitude}, \texttt{street\_name}, \texttt{landmark}, USGS State Plane 
  Coordinate System, and \texttt{block} \textbf{\#}.
  \jy{use raw names from the data for consistency.}

@@

During the 10-year period from 2014 to 2023,  NYC's 311 system
experienced a 50\% increase in the number of SRs, 
reflecting a growing public reliance on and usage of 
the 311 system. (During this 10\mbox{-}year time-frame, the 
population of NYC grew just 0.6\%.). The most significant increase
in activity occurred in 2020, when the volume of requests 
surged to 3.23 million during the COVID\mbox{-}19 pandemic, 
illustrating the essential role of the 311 system in supporting 
NYC residents during times of crisis. This steady 
rise in SRs can be at least partially attributed to the increased 
accessibility of the 311 system via online and mobile 
platforms, as well as heightened public awareness of the service. 
While the spike in 2020 was exceptional, the broader trend 
indicates sustained growth in SR volume in future years,  
highlighting the system's expanding role in managing 
both routine City operations and extraordinary events.

\jy{Remove the text above each figure. They should be included in the
  caption.}

Figure~\ref{fig:SRcountbyAgency} provides a breakdown of SRs by
the responsible Agency, showing the cumulative 
percentage of SRs handled by each
agency over the 2\mbox{-}year time-frame. Note that the
distribution of SRs is heavily concentrated among a few key 
agencies. The six largest agencies are:
\begin{enumerate}[left=1.5em]
    \item New York Police Department (NYPD): 43\% of all SRs
    \item Housing Preservation and Development (HPD): 21\%
    \item New York City Department of Sanitation (DSNY): 10\%
    \item Department of Transportation (DOT): 7\%
    \item Department of Environmental Protection (DEP): 5\%
    \item Department of Parks and Recreation (DPR): 4\%
\end{enumerate}

@@

The \textit{Mostly Empty} category includes fields such as
\texttt{taxi\_company\_borough}, \texttt{due\_date},
\jy{These variables seem only relevant to certain agencies, which
  could be made available in a separate table.}
\texttt{road\_ramp}, and \texttt{bridge\_highway\_name}.
The \textit{Partially Empty} category includes fields such as
\texttt{location\_type}, \texttt{landmark}, 
and \texttt{cross\_street\_1 \& \_2}. And the \textit{Few/None Empty} 
category includes \texttt{created\_date}, \texttt{complaint\_type},
\texttt{agency}, and \texttt{status}. For some analysis efforts, it may be
prudent to inquire further as to why some fields are 
mostly or almost always blank. Figure~\ref{fig:blank_fields} is 
a graphic depiction of total empty (blank \& N/As) for each 
field illustrating the grouping into 
the \textbf{Most}, \textbf{Partial}, and \textbf{Few/None} categories.

@@

\begin{itemize}[left=1.5em]
    \item SRs with a \texttt{closed\_date} that occurs before the 
    \texttt{created\_date} resulting a nonsensical ``negative duration''. (12,450)
\jy{what's the number in the parentheses?}
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} in 
    the far distant past, e.g. ``1900-01-01''. (8)
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} that 
    match \textbf{to the second} creating an impossible ``zero duration''. (163,720)
    \item An unusually large number of SRs closed and/or created 
    exactly at midnight, to the second. (247,000)
\end{itemize}

@@

\begin{table}[tbp]
  \centering
  \caption{Sample  Negative Durations [excluding extreme values that are <$-$730 days]}
   \label{tab:largest-errors}
 	\begin{tabular}{l l l r l}
    \toprule
    \texttt{created\_date} & \texttt{closed\_date} & Duration (days)}  & {Agency} \\
    \midrule
    2023-01-27 14:40:00 & 2022-01-14 14:40:00 & $-$378 & DOT \\
    2023-01-18 10:06:00 & 2022-01-12 10:06:00 & $-$371 & DOT \\
    2023-01-27 14:36:00 & 2022-01-22 14:35:00 & $-$370 & DOT \\
    2023-01-11 11:10:00 & 2022-01-09 11:10:00 & $-$367 & DOT \\
    2023-12-18 03:13:00 & 2023-01-16 13:10:00 & $-$335 & DOT \\
    \bottomrule
    \end{tabular}
 \end{table}

\paragraph{Identical Created \& Closed Dates (Zero Durations):}
A more prevalent issue occurs when the \texttt{closed\_date} and 
\texttt{created\_date} have exactly the same date\mbox{-}time stamp, 
to the second. This creates a \emph{zero duration} for the 
SR, which is nonsensical. There are 163,720 such zero duration 
SRs, representing 2.6\% of all non\mbox{-}blank data. 
Additionally, 99\% percent of these zero duration SRs 
occur within five agencies: Department of Health \& Mental 
Hygiene (DHMH), DOT, Department of Buildings (DOB), DSNY, and 
DEP. This distribution \textbf{does not} 
mirror the overall SR distribution by agency, suggesting 
an Agency\mbox{-}specific issue.
